PEACE: {\underline P}arallel {\underline E}nvironment for {\underline A}ssembly and {\underline C}lustering of Gene {\underline E}xpression

### 2010-05-11 papers/NAR_web_server_2009/paper.tex
Manuscript update: final(?) version

Clustering is a computationally challenging problem; the runtime and
memory requirements to cluster on the basis of pair-wise sequence
alignments make such an approach infeasible in practice.  To deal with
this, \peace\/ combines our own version of the $d^2$ alignment-free
sequence distance function \cite{Hide94} and the concept of a {\it
  minimum spanning tree} \cite{Prim57} to quickly and accurately find
clusters of ESTs expressed from the same gene without reference to a
sequenced genome.  Compared against clustering tool in the literature
\cite{Burke99,Slater00,Huang99,Parkinson02,Kalyanaraman03,Malde03,
  Pertea2003,Ptitsyn05,Hazelhurst08a,Picardi09}, \peace\/ proves to be
of equivalent quality to the \wcd\/ and \tgicl tools
\cite{Hazelhurst08a,Pertea2003}, and more sensitive and robust than
other tools.  From the user perspective, o tool in the literature can
match \peace for ease of instillation, use, or the post-clustering
analysis tools \peace provides.

@@

The \peace\/ GUI can be launched by through the
Jar file available on the \peace\/ website
(\href{http://www.peace-tools.org}{www.peace-tools.org}) on any
machine running the standard Java Virtual Machine (JVM).  Once
running, the user can employ the GUI to install the clustering engine
and perform a clustering of a data file in {\sc fasta} format, view an
initial analysis of the clusters, and produce files containing subsets
of the clusters as input to assembly tools such as \capthree\/ 
\cite{Huang99}.  A typical (first) use of \peace\/ must be
performed in the following manner (see Figure~\ref{fig:workflow}):

@@

\mc{In the course of our simulations we also investigated the memory
  footprint -- of particular concern in the large data sets commonly
  encountered.  In short, we find that, as with \wcd, the required
  memory for an execution is linear in the size of the data set --
  requiring slightly more memory than \wcd\/ and considerably less
  than \capthree.  We present relevant results in Section D.4 and
  Figure~S3 of the Supplementary Materials.}

% Appologizes to anyone working with this: the coloring would not
% presist between columns, and the break point kept changing on me.
% Modifying the \mc macro will allow for color removal without having
% to take all of this out.
\mc{We note, as a point of interest, that we can significantly improve
  the quality of \peace\/} \mc{simulation} \mc{results} \mc{through} \mc{the} \mc{increase} \mc{of}
  \mc{the} \mc{threshold} \mc{value} \mc{(see} \mc{Supplementary} \mc{Materials,} \mc{Section~D.5) --}
  \mc{achieving} \mc{a} \mc{significant} \mc{improvement} \mc{in PEACE sensitivity without an
  adverse effect on the Jaccard Index.  However, the improvement does
  not carry over to the application of real data (Supplementary
  Materials, Section E.2), where we observe a significant increase in
  the incorrect merging of clusters.  While this might be acceptable
  to a user planning to employ an assembly tool capable of breaking
  up the clusters (but unable to re-merge a split cluster), for the
  purposes of clustering assessment it makes sense base our analysis (and
  set out default values) at the lower threshold.}

\begin{figure}
  \centerline{\includegraphics[width=3.35in]{pics.d/SeJiT1T2x40.png}}
  \caption{Comparisons of Sensitivity, Jaccard Index, Type 1 error and
    Type 1 error, based on the average over 30 simulated Sanger
    Sequence ESTs sets derived from 100 zebra fish genes (see
    Supplementary Materials, Section D, for more details).  Blue/Solid
    = \peace, Green/Dash = \wcd\/ \mc{(version 0.5.1)}, Black/Dot-Dash =
    \capthree; vertical tics = 95\% confidence intervals on estimates.
    Intervals are not presented for Type 1 error due to the effective
    lack of variance.}\label{SeJiT1T2}
\end{figure}

\begin{figure}
   \centerline{\includegraphics[width=3.35in]{pics.d/seq_parallel.png}}
   \caption{Comparisons of Runtime: On left, we compare the sequential
     runtime of \peace\/ (blue) and \wcd\/ (green) on simulate sets
     ranging in size from 10K sequences to 100K sequences.  (\capthree\/
     runtime is not reported, as the time spent on clustering cannot
     be differentiated from the time spent on assembly.)  On right,
     a comparison when run in parallel, ranging the number of
     processors from 2 to 12 while holding the EST/Processor ratio
     steady at the constant 8888.  All values represent the average
     of 30 runs; vertical tics = 95\% confidence intervals on
     estimates.  All runs were done on a 3.0 GHz Intel Xeon EM64T CPU with a 2 MB
     cache and a 800 MHz front side bus, model number Xeon LV 3.0 (2005).}\label{seq_parallel}
\end{figure}

While \peace\/ clearly matches \wcd\/ on long-read sequence, comparisons
against \tgicl\/ \cite{Pertea2003}  are more complicated.  In short, we find that when
parameters are picked appropriately, \tgicl\/ is more sensitive at lower
error rate -- but \peace\/ is more robust to error, in all cases
has an improved Type 2 error rates and is better able to
distinguish duplicated genes.  (See the Supplementary Materials for a
more detailed analysis.)

In applying the tools to real data Sanger data, we used the Human
Benchmark Data Set used to test \easycluster\/ \cite{Picardi09}, and
the A076941 {\it Arabidopsis thaliana} data set used to test wcd\/
\cite{Hazelhurst08a,Hazelhurst08b} (see Table~\ref{SangerReal}).  We
notice essentially identical results for \peace\/ and wcd\/ in
quality, both significantly better than \capthree\/ in Sensitivity and
Type 1 error rate, while slightly worse in the Jaccard Index and Type
2 error.  In runtime we see some inconsistency, with \peace\/ showing
a 60\% runtime improvement over \wcd\/ in the first data set, but
requiring 20\% more time than \wcd\/ in the Arabidopsis data set.  (We
cannot make a reasonable runtime comparison to \capthree, as the
runtime of that tool reflects both clustering and assembly.)  In
Table~S1 of the Supplementary Materials we present runtimes of several
more sets, observing that while \peace\/ appears to be significantly
faster on the smaller sets, \wcd\/ does overtake it for larger sets.

\begin{table*}[tbp]
\begin{center}
\begin{tabular} {| c || c || c | c | c | c | c | c | c |}
\hline
 & 
& Sensitivity & Jaccard & \begin{tabular}{c} Type 1 \\
  error \end{tabular} & \begin{tabular}{c} Type 2 \\
  error \end{tabular} &
\begin{tabular}{c} Number \\ of  \\
    Clusters \end{tabular} & \begin{tabular}{c} Number \\ of \\
    Singletons \end{tabular}
    & \begin{tabular}{c} Single \\ processor \\ runtime (s) \end{tabular}\\
\hline
\hline EasyCluster Human & \peace & 0.998 & 0.672 & 0.153 & 0.042 & 118 & 21 & 293 \\
\cline{2-9} Benchmark & \wcd & 0.998 & 0.672 & 0.144 & 0.044 & 113 & 16 & 804 \\
\cline{2-9} (111 Genes) & \capthree & 0.657 & 0.643 & 1.000 & 0.001 & 2269 & 1827 & \mbox{NA} \\
\hline
\hline WCD A076941 & \peace & 0.932 & 0.475 & 0.351 & 0.027 & 18825 & 8951 & 1166\\
\cline{2-9} Benchmark & \wcd & 0.933 & 0.476 & 0.350 &  0.027 & 18787 & 8553 & 966 \\
\cline{2-9} (13240 genes) & \capthree & 0.826 & 0.802 & 0.486 & 0.014 & 25042 & 14916 &
\mbox{NA} \\
\hline
\end{tabular}
\end{center}
\caption{Comparisons of runs on the EasyCluster human Benchmark
  Data Set and the WCD A076941 Arabidopsis thaliana data set using the
  standard quality measurements.}\label{SangerReal}
\end{table*}

@@

We tested the three tools on short-read data using the \metasim\/ tool
of {\it Richter et al.} \cite{Richter2008}.  Encoded into \metasim\/
are sequence generation and error models corresponding to several
technologies, including the 454 short-read technology (producing reads
ranging in size from 200bp to 370bp with a mean of 250bp and a
standard deviation of 17bp), and the Illumina short-read technology
(producing reads of exactly 62bp).  In the 454 experiments \peace\/
achieved a sensitivity of 0.871, an 89.7\% improvement in sensitivity
over \wcd\/ and a 193\% improvement in sensitivity over \tgicl, with
comparable values for the Jaccard Index.  \wcd\/ was unable to handle
the Illumina data, while \peace\/ sensitivity, at a value of 0.384,
was nearly twice as good as Illumina.  (For more details, see Table~S3
in the Supplementary Materials.)  In short, \peace\/ performs quite
well for 454 reads and provides some useful information for Illumina
reads, with considerably better results than \wcd\/ or \tgicl\/.

@@

\mc{Here we have presented \peace, a stand alone tool for the
  high-throughput clustering of transcript fragments capable of
  dealing with sequences as short as 50 bases.  \peace\/ is
  open-source and managed through a user friendly GUI that enables
  both local and remote installation and execution in sequential or
  parallel mode.  Based on a novel minimum spanning tree based
  algorithm for the clustering of the fragments by gene association,
  \peace\/ shows significant improvement in sensitivity over the
  competing \wcd\/} \mc{tool} \cite{Hazelhurst08a} \mc{when} \mc{applied} \mc{to} \mc{NGS}
  \mc{reads,} \mc{matches} \mc{\wcd\/} \mc{when} \mc{applied} \mc{to} \mc{Sanger} \mc{sequencing}
  \mc{output,} \mc{and} \mc{shows} \mc{an} \mc{order} \mc{of} \mc{magnitude} \mc{in improvement over the
  clustering performed in the course of assembly by the \capthree\/
  tool \cite{Huang99}.}

As a clustering tool based on sequence distance, \peace\/ faces
certain inherent limitations. For example, \peace\/ cannot handle
duplicate genes; like \wcd\/, it is unable to separate clusters
corresponding to genes with a greater than 88\% similarity.
Similarly, other natural biological effects (e.g. the trans-splicing of
transcripts), effects from poorly cleaned transcript data (e.g. the
failure to remove sequencing adapters or post-transcriptional
poly(A)/(T) tails), and the presence of low-complexity repeats can
cause similar effects in these clustering tools.  These problems can be
handled through the application of the assembler, and the ability to
apply any assembler to small clusters (as opposed to the data set as a
whole) results in a significant reduction in overall assembly time.

@@

% LocalWords:  arallel nalysis lustering ngine Rao Moler Ozden Zhang Liang ESTs
% LocalWords:  Karro XXXXX transcriptome pre isoforms SNPs unsequenced runtime
% LocalWords:  chlamydomondomonas ESTate PaCE xsact CLU WCD Hazelhurst et al th
% LocalWords:  CHUN nalyzer Prim's MSTs subgraph MPI SourceCode ESTSim runtimes
% LocalWords:  ests Chlamydomonas Reihartdii gmap CDNA transcriptomes cDNA www
% LocalWords:  reinhartdii conifergdb DNAc bp wcd reinhardtii PhRMA Iddo fasta
% LocalWords:  Friedberg Scoville OpenMPI initio multi Screenshots nvironment
% LocalWords:  ssembly xpression GenBank dbEST pyrosequencing ESTsim sim mRNAs
% LocalWords:  NGS polyA set's GHz Xeon Bikun Arabidopsis JVM Jaccard LV GUI's
% LocalWords:  Data Set thaliana data set EasyCluster MetaSim Illumina

>>>>

Clustering is a computationally challenging problem; the runtime and
memory requirements to cluster on the basis of pair-wise sequence
alignments make such an approach infeasible in practice.  To deal with
this, \peace\/ combines our own version of the $d^2$ alignment-free
sequence distance function \cite{Hide94} and the concept of a {\it
  minimum spanning tree} \cite{Prim57} to quickly and accurately find
clusters of ESTs expressed from the same gene without reference to a
sequenced genome.  Compared against clustering tool in the literature
\cite{Burke99,Slater00,Huang99,Parkinson02,Kalyanaraman03,Malde03,
  Pertea2003,Ptitsyn05,Hazelhurst08a,Picardi09}, \peace\/ proves to be
produce results of quality competitive with the \wcd\/ and \tgicl\/ tools
\cite{Hazelhurst08a,Pertea2003}, and more sensitive and robust than
other tools.  From the user perspective, mo tool in the literature can
match \peace\/ for ease of instillation or use, or the post-clustering
analysis tools \peace provides.

@@

The \peace\/ GUI, available as a {\small JAR} file from the \peace\/
website (\href{http://www.peace-tools.org}{www.peace-tools.org}),
can be run on any machine supporting the standard Java Virtual
Machine (JVM).  The user can employ the GUI to install
the clustering engine and perform a clustering of a data file in {\sc
  fasta} format, view an analysis of the clusters, and produce
files containing subsets of the clusters as input to assembly tools
such as \capthree\/ \cite{Huang99}.  A typical (first) use of \peace\/
must be performed in the following manner (see
Figure~\ref{fig:workflow}):

@@

\mc{In the course of our simulations we also investigated the memory
  footprint -- of particular concern in the large data sets commonly
  encountered.  We find that, as with \wcd, the memory used for
  execution is linear in the size of the data set, requiring
  slightly more memory than \wcd\/ and considerably less than
  \capthree.  (See Section D.4 and Figure~S3 of the Supplementary
  Materials.}

% Appologizes to anyone working with this: the coloring would not
% presist between columns, and the break point kept changing on me.
% Modifying the \mc macro will allow for color removal without having
% to take all of this out.
\mc{We note, as a point of interest, that we can significantly improve
  the quality of \peace\/} \mc{simulation} \mc{results} \mc{through} \mc{the} \mc{increase} \mc{of}
  \mc{the} \mc{threshold} \mc{value} \mc{(see} \mc{Supplementary} \mc{Materials,} \mc{Section~D.5) --}
  \mc{achieving} \mc{a} \mc{significant} \mc{improvement} \mc{in PEACE sensitivity without an
  adverse effect on the Jaccard Index.  However, the improvement does
  not carry over to the application of real data (Supplementary
  Materials, Section E.2), where we observe a significant increase in
  the incorrect merging of clusters.  While this might be acceptable
  to a user planning to employ an assembly tool capable of breaking
  up the clusters (but unable to re-merge a split cluster), for the
  purposes of clustering assessment it makes sense to base our analysis (and
  set out default values) at the lower threshold.}

\begin{figure}
  \centerline{\includegraphics[width=3.35in]{pics.d/SeJiT1T2x40.png}}
  \caption{Comparisons of Sensitivity, Jaccard Index, Type 1 error and
    Type 1 error, based on the average over 30 simulated Sanger
    Sequence ESTs sets derived from 100 zebra fish genes (see
    Supplementary Materials, Section D, for more details).  Blue/Solid
    = \peace, Green/Dash = \wcd\/ \mc{(version 0.5.1)}, Black/Dot-Dash =
    \capthree; vertical tics = 95\% confidence intervals on estimates.}\label{SeJiT1T2}
\end{figure}

\begin{figure}
   \centerline{\includegraphics[width=3.35in]{pics.d/seq_parallel.png}}
   \caption{Comparisons of Runtime: On left, we compare the sequential
     runtime of \peace\/ (blue) and \wcd\/ (green) on simulate sets
     ranging in size from 10K sequences to 100K sequences.  (\capthree\/
     runtime is not reported, as the time spent on clustering cannot
     be differentiated from the time spent on assembly.)  On right,
     a comparison when run in parallel, ranging the number of
     processors from 2 to 12 while holding the EST/Processor ratio
     steady at the constant 8888.  All values represent the average
     of 30 runs; vertical tics = 95\% confidence intervals on
     estimates.  All runs were done on a 3.0 GHz Intel Xeon EM64T CPU with a 2 MB
     cache and a 800 MHz front side bus, model number E5520 (2005).}\label{seq_parallel}
\end{figure}

Comparisons against \tgicl\/ \cite{Pertea2003} are more complicated.
We find that when \tgicl\/ is more sensitive at lower error rates,
\peace\/ is more robust to error, and in all cases \peace\/ has an
improved Type 2 error rates and is better able to distinguish
duplicated genes.  (See Supplementary Materials Section D.6.)

In applying the tools to real data Sanger data, we used the Human
Benchmark Data Set used to test \easycluster\/ \cite{Picardi09}, and
the A076941 {\it Arabidopsis thaliana} data set used to test wcd\/
\cite{Hazelhurst08a} (see Table~\ref{SangerReal}).  We notice mixed
results, with \peace, \wcd, and \tgicl\/ showing comparable similarity,
\peace\/ and \wcd\/ showing some superiority to \tgicl\/ in Type 1 error
rate but with a poorer showing in Type 2 error, and with all three
tools coming considerably closer to the correct number of clusters
than does \capthree.  In runtime we see some inconsistency, with
\peace\/ showing a 60\% runtime improvement over \wcd\/ in the first
data set and a time comparable to \tgicl, but requiring 20\% more time
than \wcd\/ in the Arabidopsis data set and losing to \tgicl\/ by a
large margin.  In Table~S1 of the Supplementary Materials we present
runtimes of several more sets for the distance-based tool, observing
that while \peace\/ appears to be significantly faster on the smaller
sets, \wcd\/ does overtake it for larger sets.

\begin{table*}[tbp]
\begin{center}
\begin{tabular} {| c || c || c | c | c | c | c | c | c |}
\hline
 & 
& Sensitivity & Jaccard & \begin{tabular}{c} Type 1 \\
  error \end{tabular} & \begin{tabular}{c} Type 2 \\
  error \end{tabular} &
\begin{tabular}{c} Number \\ of  \\
    Clusters \end{tabular} & \begin{tabular}{c} Number \\ of \\
    Singletons \end{tabular}
    & \begin{tabular}{c} Single \\ processor \\ runtime (s) \end{tabular}\\
\hline
\hline EasyCluster Human & \peace   & 0.998 & 0.672 & 0.153 & 0.042 & 118 & 21 & 293 \\
\cline{2-9} Benchmark & \wcd        & 0.998 & 0.672 & 0.144 & 0.044 & 113 & 16 & 804 \\
\cline{2-9} (111 Genes) & \capthree & 0.657 & 0.643 & 1.000 & 0.001 & 2269 & 1827 & \mbox{NA} \\
\cline{2-9}             & \tgicl    & 0.998 & 0.949 & 0.568 & 0.018 & 221 & 86 & 278 \\
\hline
\hline A076941 & \peace           & 0.932 & 0.475 & 0.351 & 0.027 & 18825 & 8951 & 1166\\
\cline{2-9} Benchmark & \wcd          & 0.933 & 0.476 & 0.350 & 0.027 & 18787 & 8553 & 966 \\
\cline{2-9} (13240 genes) & \capthree & 0.826 & 0.802 & 0.486 & 0.014 & 25042 & 14916 & \mbox{NA} \\
\cline{2-9} & \tgicl                  & 0.939 & 0.209 & 0.401 & 0.020 & 20248 & 1065 & 425 \\
\hline
\end{tabular}
\end{center}
\caption{Comparisons of runs on the EasyCluster Human Benchmark
  Data Set and the WCD A076941 Arabidopsis thaliana data set using the
  standard quality measurements.}\label{SangerReal}
\end{table*}

@@

We tested the tools on short-read data using the \metasim\/ tool
of {\it Richter et al.} \cite{Richter2008}.  Encoded into \metasim\/
are sequence generation and error models corresponding to several
technologies, including the 454 short-read technology (producing reads
ranging in size from 200bp to 370bp with a mean of 250bp and a
standard deviation of 17bp), and the Illumina short-read technology
(producing reads of exactly 62bp).  In the 454 experiments \peace\/
achieved a sensitivity of 0.871, an 89.7\% improvement in sensitivity
over \wcd\/ and a 193\% improvement in sensitivity over \tgicl, with
comparable values for the Jaccard Index.  \wcd\/ was unable to handle
the Illumina data, while \peace\/ sensitivity, at a value of 0.384,
was nearly twice as good as TGICL.  (For more details, see Table~S3
in the Supplementary Materials.)  In short, \peace\/ performs quite
well for 454 reads and provides some useful information for Illumina
reads, with considerably better results than \wcd\/ or \tgicl\/.

@@

Here we have presented \peace, a stand alone tool for the
high-throughput clustering of transcript fragments capable of dealing
with sequences as short as 50 bases.  \peace\/ is open-source and
managed through a user friendly GUI that enables both local and remote
installation and execution in parallel mode.  Using a novel minimum
spanning tree based algorithm for the clustering of fragments by gene
association, \peace\/ shows significant improvement in sensitivity
over the competing \wcd\/ tool and \tgicl tool when applied to NGS
reads \cite{Hazelhurst08a,Pertea2003}, matches \wcd\/ when applied to
Sanger sequencing output, and shows an order of magnitude in
improvement over the clustering performed in the course of assembly by
the \capthree\/ tool \cite{Huang99}.

As a clustering tool based on sequence distance, \peace\/ faces
certain inherent limitations. \peace\/ cannot handle duplicate genes;
like \wcd\/, it is unable to separate clusters corresponding to genes
with a greater than 88\% similarity.  Similarly, other natural
biological effects (e.g. the trans-splicing of transcripts), effects
from poorly cleaned transcript data (e.g. the failure to remove
sequencing adapters or post-transcriptional poly(A)/(T) tails), and
the presence of low-complexity repeats can cause similar effects in
these clustering tools.  These problems can be handled through the
application of the assembler, and the ability to apply any assembler
to small clusters results in a significant reduction in overall
assembly time.

@@

% LocalWords:  arallel nalysis lustering ngine Rao Moler Ozden Zhang Liang ESTs
% LocalWords:  Karro XXXXX transcriptome pre isoforms SNPs unsequenced runtime
% LocalWords:  chlamydomondomonas ESTate PaCE xsact CLU WCD Hazelhurst et al th
% LocalWords:  CHUN nalyzer Prim's MSTs subgraph MPI SourceCode ESTSim runtimes
% LocalWords:  ests Chlamydomonas Reihartdii gmap CDNA transcriptomes cDNA www
% LocalWords:  reinhartdii conifergdb DNAc bp wcd reinhardtii PhRMA Iddo fasta
% LocalWords:  Friedberg Scoville OpenMPI initio multi Screenshots nvironment
% LocalWords:  ssembly xpression GenBank dbEST pyrosequencing ESTsim sim mRNAs
% LocalWords:  NGS polyA set's GHz Xeon Bikun Arabidopsis JVM Jaccard LV GUI's
% LocalWords:  Data Set thaliana data set EasyCluster MetaSim Illumina tgicl
